---
title: "Frequently asked questions for the sommer package"
author: "Giovanny Covarrubias-Pazaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ for the sommer package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The sommer package was developed to provide R users a powerful and reliable multivariate mixed model solver. The package is focused in problems of the type p > n (more effects to estimate than observations) and its core algorithm is coded in C++ using the Armadillo library. This package allows the user to fit mixed models with the advantage of specifying the variance-covariance structure for the random effects, and specify heterogeneous variances, and obtain other parameters such as BLUPs, BLUEs, residuals, fitted values, variances for fixed and random effects, etc. 

The purpose of this vignette is to provide answers to frequently asked questions (FAQ) related to performance and possible issues:  

### 1) I got an error similar to: 

```{r}
# iteration    LogLik     wall    cpu(sec)   restrained
#     1      -224.676   18:11:23      3           0
# Sistem is singular. Stopping the job
# matrix multiplication: incompatible matrix dimensions: 0x0 and ...x...
```

This error indicates that your model is singular (phenotypic variance V matrix is not invertible) and therefore the model is stopped throwing the "incompatible matrix dimensions: 0x0 and ...x..." error message. Whether you can try a simpler model or just modify the argument `tolparinv` in the `mmer` function. The default is 1e-6, which means that it will try to invert V and if it fails it will try to add a small value to the diagonal of V of 1e-6 to make it invertible. If this fails then the program will stop returning that error message which should make you check the quality of your data or model attempted.

Sometimes the model becomes singular when you use variance covariance matrices (i.e. genomic relationship matrices) that are not full-rank. You can try to make it full-rank and try again.

## 2) My model runs very slow

Keep in mind that sommer uses direct inversion (DI) algorithm which can be very slow for large datasets. The package is focused in problems of the type p > n (more random effect levels than observations) and models with dense covariance structures. For example, for experiment with dense covariance structures with low-replication (i.e. 2000 records from 1000 individuals replicated  twice with a covariance structure of 1000x1000) sommer will be faster than MME-based software. Also for genomic problems with large number of random effect levels, i.e. 300 individuals (n) with 100,000 genetic markers (p). For highly replicated trials with small covariance structures or n > p (i.e. 2000 records from 200 individuals replicated 10 times with covariance structure of 200x200) asreml or other MME-based algorithms will be much faster and we recommend you to opt for those software.

## 3) Can I run rrBLUP for markers and GBLUP for individuals in sommer?

Both types of models can be fitted in sommer. The only thing that it changes is what is the random effect of interest; the marker matrix or the identifier for the individual. Here there is a complex example using multi-trait models but can be used with only one trait.

```{r}
library(sommer)
## rrBLUP for makers
data(DT_cpdata)
mix.rrblup <- mmer(fixed=color~1,
                   random=~vs(GT,Gtc=unsm(1)) + vs(Rowf,Gtc=diag(1)),
                   rcov=~vs(units,Gtc=unsm(1)), getPEV = FALSE,
                   data=DT, verbose = FALSE)
summary(mix.rrblup)
## GBLUP for individuals
A <- A.mat(GT)
mix.gblup <- mmer(fixed=color~1,
                  random=~vs(id,Gu=A, Gtc=unsm(1)) + vs(Rowf,Gtc=diag(1)),
                  rcov=~vs(units,Gtc=unsm(1)),
                  data=DT, verbose = FALSE)
summary(mix.gblup)
## Equivalence
plot(GT%*%mix.rrblup$U$`u:GT`$color, mix.gblup$U$`u:id`$color)
```

## 4) I am missing BLUPs for individuals even when I provided them in the relationship matrix

I got this good question in the past: "when I want to fit an animal model with sommer package using additive relationship matrix(A), this A matrix would contain parents. But the random effects only contains animals in the random effect but not including parents in the A matrix. How can I get the random effects for parents?""

Answer: The easy way to do it is to make sure that even if the parents don't show up in the dataset, you need to make sure that they are present in the levels of the column that contains the individuals (i.e. animal IDs), in addition they have to be provided in the relationship matrix and that's it. They should be returned in the blups.

```{r}
library(sommer)

data(DT_cpdata)
#### create the variance-covariance matrix
A <- A.mat(GT) # additive relationship matrix
#### look at the data and fit the model
set.seed(12)
DT2 <- droplevels(DT[sample(1:nrow(DT),100),]) # we simulate a dataset with only 100 animals
nrow(DT2); length(levels(DT2$id))

# we fit a model with the reduced datatset where only 100 blups will be returned since only
# 100 levels exist in the "id" column
mix1 <- mmer(Yield~1,
              random=~vs(id,Gu=A)
                      + Rowf + Colf,
              rcov=~units,
              data=DT2, verbose = FALSE)
summary(mix1)
length(mix1$U$`u:id`$Yield) # only 100 levels

# we add additional levels to the "id" column and also provide them in the relationship matrix
levels(DT2$id) <- c(levels(DT2$id), setdiff(levels(DT$id), levels(DT2$id)))
mix2 <- mmer(Yield~1,
             random=~vs(id,Gu=A)
             + Rowf + Colf,
             rcov=~units,
             data=DT2, verbose = FALSE)
summary(mix2)
length(mix2$U$`u:id`$Yield) # now 363 levels
```

## 5) How can I use the AR1(), CS() and ARMA() functions

Sommer doesn't support the estimation of additional correlation components like AR1 in the way asreml does. Still, if the user know the correlation value or can do an iterative approach to find the best value then these functions can be used to specify the variance covariance structure for a given random effect.

For example, in the DT_cpdata dataset we have a field with row and column coordinates. This allows to fit row and column as random effects:

```{r}
library(sommer)
data(DT_cpdata)
mix1 <- mmer(Yield~1,
              random=~ Rowf + Colf,
              rcov=~units,
              data=DT, verbose = FALSE)
summary(mix1)$varcomp
```

If the user wants to relax the independence between rows and define an AR1 covariance structure among rows then the model could be fitted as:

```{r}
library(sommer)
data(DT_cpdata)
mixAR1row <- mmer(Yield~1,
             random=~ vs(Rowf, Gu=AR1(Rowf, rho=0.3)) + Colf,
             rcov=~units,
             data=DT, verbose = FALSE)
summary(mixAR1row)$varcomp
```

Same could be done for the column random effect:

```{r}
library(sommer)
data(DT_cpdata)
mixAR1col <- mmer(Yield~1,
             random=~ Rowf + vs(Colf, Gu=AR1(Colf, rho=0.3)),
             rcov=~units,
             data=DT, verbose = FALSE)
summary(mixAR1col)$varcomp
```

If on the other hand, you would like to stablish the presence of correlation in row and columns at the same time the model would look like this:

```{r}
library(sommer)
data(DT_cpdata)
mixAR1rowcol <- mmer(Yield~1,
                  random=~ vs(Rowf:Colf, 
                              Gu=kronecker(AR1(Rowf, rho=0.3),AR1(Colf, rho=0.3),make.dimnames = TRUE) 
                              ),
                  rcov=~units,
                  data=DT, verbose = FALSE)
summary(mixAR1rowcol)$varcomp
```

Notice that if you specify a random effect that is the interaction between 2 random effects the covariance structure to be specified in the `Gu` argument has to be built using the `kronecker()` function. The same applies to the `ARMA()` and `CS()` functions. Please keep in mind that the correlation values (rho argument) is a fixed value not estimated by REML like asreml does but you can always follow an iterative approach.

## 6) Can I run GWAS in MET experiments with replicates? 

Although the direct-inversion algorithm that sommer uses in the background is not the best choice to solve GWAS models for the n > p scenario it is still possible to perform GWAS in MET models.

For example, assume a MET model that has 41 lines with 1000 SNP markers, tested in 3 environments. The genetic term in a MET can be modeled as CS, DIAG or US covariance. The whole point of the GWAS function in sommer is to provide the name of the gTerm (random effect) that will match the marker data provided. 

We first make up marker data for the MET data for example purposes
```{r}
library(sommer)
data(DT_example)
M <- matrix(rep(0,41*1000),1000,41)
for (i in 1:41) {
  M[,i] <- ifelse(runif(1000)<0.5,-1,1)
}
tM <- t(M)
```

We then fit the MET model of the type compound simmetry (CS) and evaluate the GWAS for the main genetic term "Name":

```{r}
## GWAS for main term in CS model
ansx <- GWAS(Yield~Env,
             random= ~ Name + Env:Name,
             rcov= ~ units,
             data=DT,
             M=tM,
             gTerm = "Name", verbose = FALSE)

ms <- as.data.frame(t(ansx$scores))
plot(ms$`Yield score`, ylim=c(0,8))
```

We could do the same but evaluate the GWAS considering the genetic term as the interaction term "Env:Name"

```{r}
## GWAS for the interaction term in CS model
E <- matrix(1,nrow = length(unique(DT$Env)));E
EtM <- kronecker(E,tM)
ansx <- GWAS(Yield~Env,
             random= ~ Name + Env:Name,
             rcov= ~ units,
             data=DT,
             M=EtM,
             gTerm = "Env:Name", verbose = FALSE)

ms <- as.data.frame(t(ansx$scores))
plot(ms$`Yield score`, ylim=c(0,8))
```

If the MET is a diagonal model, there's a variance components and BLUPs for each environment. Therefore we can evaluate the GWAS at any environment, here for example we evaluate the GWAS at the genetic term in the environment 'CA.2011'.

```{r}
## GWAS for the interaction term in DIAG model
E <- matrix(1,nrow = length(unique(DT$Env)));E
EtM <- kronecker(E,tM)
ansx <- GWAS(Yield~Env,
             random= ~Name + vs(ds(Env),Name),
             rcov= ~ vs(ds(Env),units),
             data=DT,
             M=tM,
             gTerm = "CA.2011:Name", verbose = FALSE )

ms <- as.data.frame(t(ansx$scores))
plot(ms$`Yield score`, ylim=c(0,8))
```

If the MET is an unstructured model, there's a variance components and BLUPs for each environment and a covariance component among the different combinations of environments. Therefore we can evaluate the GWAS at any environment as before, here for example we evaluate the GWAS at the genetic term in the environment 'CA.2011'. The difference with the previous model is that here we expect a greater accuracy in the environment CA.2011 since it has borrowed information from the other environments given the covariance fitted among environments.

```{r}
## GWAS for main term in US model
ansx <- GWAS(Yield~Env,
             random= ~vs(us(Env),Name),
             rcov= ~ vs(us(Env),units),
             data=DT,
             M=tM,
             gTerm = "CA.2011:Name", verbose = FALSE)

ms <- as.data.frame(t(ansx$scores))
plot(ms$`Yield score`, ylim=c(0,8))
```

Same theory applies for the multitrait model:

```{r}
## GWAS for main term in multitrait DIAG model
ansx <- GWAS(cbind(Weight,Yield)~Env,
             random= ~vs(ds(Env),Name, Gtc=unsm(2)),
             rcov= ~ vs(ds(Env),units, Gtc=diag(2)),
             data=DT,
             M=tM,
             gTerm = "CA.2011:Name", verbose = FALSE)

ms <- as.data.frame(t(ansx$scores))
plot(ms$`Yield score`, ylim=c(0,8))
```


## Literature

Covarrubias-Pazaran G. 2016. Genome assisted prediction of quantitative traits using the R package sommer. PLoS ONE 11(6):1-15.

Covarrubias-Pazaran G. 2018. Software update: Moving the R package sommer to multivariate mixed models for genome-assisted prediction. doi: https://doi.org/10.1101/354639

Bernardo Rex. 2010. Breeding for quantitative traits in plants. Second edition. Stemma Press. 390 pp.

Gilmour et al. 1995. Average Information REML: An efficient algorithm for variance parameter estimation in linear mixed models. Biometrics 51(4):1440-1450.

Henderson C.R. 1975. Best Linear Unbiased Estimation and Prediction under a Selection Model. Biometrics vol. 31(2):423-447.

Kang et al. 2008. Efficient control of population structure in model organism association mapping. Genetics 178:1709-1723.

Lee, D.-J., Durban, M., and Eilers, P.H.C. (2013). Efficient two-dimensional smoothing with P-spline ANOVA mixed models and nested bases. Computational Statistics and Data Analysis, 61, 22 - 37.

Lee et al. 2015. MTG2: An efficient algorithm for multivariate linear mixed model analysis based on genomic information. Cold Spring Harbor. doi: http://dx.doi.org/10.1101/027201.

Maier et al. 2015. Joint analysis of psychiatric disorders increases accuracy of risk prediction for schizophrenia, bipolar disorder, and major depressive disorder. Am J Hum Genet; 96(2):283-294.

Rodriguez-Alvarez, Maria Xose, et al. Correcting for spatial heterogeneity in plant breeding experiments with P-splines. Spatial Statistics 23 (2018): 52-71.

Searle. 1993. Applying the EM algorithm to calculating ML and REML estimates of variance components. Paper invited for the 1993 American Statistical Association Meeting, San Francisco.

Yu et al. 2006. A unified mixed-model method for association mapping that accounts for multiple levels of relatedness. Genetics 38:203-208.

Abdollahi Arpanahi R, Morota G, Valente BD, Kranis A, Rosa GJM, Gianola D. 2015. Assessment of bagging GBLUP for whole genome prediction of broiler chicken traits. Journal of Animal Breeding and Genetics 132:218-228.

Tunnicliffe W. 1989. On the use of marginal likelihood in time series model estimation. JRSS 51(1):15-27.