---
title: "Installing and configuring BLAS thread control"
author: "sommer development team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing and configuring BLAS thread control}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# What the error means

An explicit request such as `mmes(..., blasThreads="auto")` can produce:

```text
Error: Runtime BLAS control is unavailable. Install flexiblas for a FlexiBLAS R installation, or RhpcBLASctl for a supported BLAS; alternatively configure BLAS before launching R and leave blasThreads=NULL.
```

This is a thread-control availability error, not an error in the model formula
or covariance structures. Three separate components are involved:

1. **The BLAS library used by R.** A threaded implementation, such as OpenBLAS
   or Intel MKL, performs dense numerical work. FlexiBLAS is a wrapper that
   selects a BLAS backend; it is not itself a guarantee of a threaded backend.
2. **An R package that controls that library.** The current `sommer`
   implementation uses `flexiblas` for an available FlexiBLAS installation,
   or `RhpcBLASctl` for a recognized supported BLAS installation.
3. **Allocation-aware CPU detection.** `parallelly` is needed specifically for
   `blasThreads="auto"`. An explicit positive integer does not require it.

These packages are optional dependencies. Installing `sommer` does not
necessarily install them, and installing a control package does not replace
R's BLAS library. No separate CHOLMOD installation is needed for
`solver="cholmod"`; that solver uses the CHOLMOD bundled with R's Matrix package.

# Identify the R installation and BLAS

Run these commands in the same R session that you use to fit the model:

```{r identify-runtime}
R.version.string
R.home()
.libPaths()
extSoftVersion()["BLAS"]
sessionInfo()[c("BLAS", "LAPACK")]
```

Look for `flexiblas`, `openblas`, or `mkl` in the BLAS path or description.
On a cluster, a different R module, terminal, RStudio session, or batch job
can use a different R installation and package library. Installing the
adapter into one installation does not make it available to another.

The current implementation recognizes non-FlexiBLAS paths containing
`openblas`, `mkl`, `acml`, or `goto` for the `RhpcBLASctl` route. A generic
name such as `libRblas.so` or `Rblas.dll` may conceal the actual backend;
the current detection does not infer its identity from that name. It also
does not currently recognize a standalone BLIS or Apple Accelerate backend
through this route. A control package's capabilities alone do not override
these detection rules.

# Install the appropriate R packages

## Route A: R uses FlexiBLAS

In the R installation used for fitting, run:

```{r install-flexiblas, eval=FALSE}
install.packages(c("flexiblas", "parallelly"),
                 repos="https://cloud.r-project.org")
```

Restart R after installation, then check:

```{r verify-flexiblas, eval=FALSE}
library(flexiblas)
stopifnot(flexiblas_avail())
flexiblas_current_backend()
flexiblas_get_num_threads()
parallelly::availableCores()
```

`flexiblas_avail()` must be `TRUE`. If it is `FALSE`, loading this R package
has not made the running R installation use FlexiBLAS. Use the correct
R installation or ask your administrator for an R build linked to FlexiBLAS.

If FlexiBLAS is available but its selected backend is a serial reference
implementation, list the available backends:

```{r select-flexiblas-backend, eval=FALSE}
flexiblas::flexiblas_list()
flexiblas::flexiblas_list_loaded()

# Only use OPENBLAS if that name appears in the available backend list.
backend <- flexiblas::flexiblas_load_backend("OPENBLAS")
flexiblas::flexiblas_switch(backend)
flexiblas::flexiblas_current_backend()
flexiblas::flexiblas_get_num_threads()
```

Backend names and availability depend on the system configuration. Switching
backends changes BLAS/LAPACK for the whole R process; do it before fitting
models and record the choice for reproducibility. `mmes()` changes the thread
setting when requested, but does not choose or install a FlexiBLAS backend.

For a FlexiBLAS R installation, installing only `RhpcBLASctl` is not a
substitute for this route: the current `sommer` controller deliberately
requires the FlexiBLAS interface when the reported BLAS is FlexiBLAS.

## Route B: R uses a recognized OpenBLAS or MKL library directly

Run:

```{r install-blasctl, eval=FALSE}
install.packages(c("RhpcBLASctl", "parallelly"),
                 repos="https://cloud.r-project.org")
```

Restart R, then check:

```{r verify-blasctl, eval=FALSE}
library(RhpcBLASctl)
extSoftVersion()["BLAS"]
blas_get_num_procs()
parallelly::availableCores()
```

The getter alone does not prove that the backend supports changing its thread
count. Some unsupported configurations return one. The small `mmes()` example
below checks the complete path; an explicit request that the backend does not
accept results in an error rather than a silent fallback.

## If R still uses a reference BLAS

Installing `RhpcBLASctl`, `flexiblas`, or an operating-system OpenBLAS package
does not by itself guarantee that the running R process uses a threaded BLAS.
R must actually load or be linked to that library.

On Linux, use a distribution-supported R/BLAS configuration or an administrator
provided R module linked to OpenBLAS, MKL, or FlexiBLAS. Some distributions use
a BLAS alternatives mechanism, while other installations require a different
R build. Follow the instructions for your distribution and R installation;
there is no universal library-switch command. For source builds, consult the
[R Installation and Administration manual](https://cran.r-project.org/doc/manuals/r-release/R-admin.html#BLAS).

On Windows, `RhpcBLASctl` may be available as a binary, but installing it does
not replace the BLAS shipped with R. Source packages need the Rtools version
matching your R release. On macOS, source packages need the applicable command
line compiler tools and, when required, a compatible Fortran compiler. Use a
supported R distribution or build with a known threaded BLAS if the current
backend cannot be controlled by this implementation. Do not manually overwrite
R's BLAS/LAPACK files with an arbitrary downloaded library.

If installation fails with missing headers or libraries, install the development
files for the already selected numerical stack, or request them from your
administrator. On a managed cluster, using its matching compiler and R modules
is preferable to mixing incompatible system and user-installed libraries.

# Verify control with a small model

First confirm that this session is loading a `sommer` build containing the
argument. Older installed builds can coexist with the current source folder:

```{r verify-sommer, eval=FALSE}
library(sommer)
packageVersion("sommer")
find.package("sommer")
stopifnot("blasThreads" %in% names(formals(mmes)))
```

The following example runs only when the required controller and
`parallelly` are available. It uses small synthetic data, rather than requiring
your `DT` or `MMTi` objects. It tests a one-thread request first, then
allocation-aware automatic sizing. Both are short verification fits, not
convergence or performance benchmarks.

```{r verification-fit}
blas <- unname(extSoftVersion()["BLAS"])
hasFlexiBLAS <- requireNamespace("flexiblas", quietly=TRUE) &&
  isTRUE(tryCatch(flexiblas::flexiblas_avail(), error=function(error) FALSE))
hasDirectControl <- !is.na(blas) &&
  !grepl("flexiblas", blas, ignore.case=TRUE) &&
  grepl("openblas|mkl|acml|goto", blas, ignore.case=TRUE) &&
  requireNamespace("RhpcBLASctl", quietly=TRUE)

if(requireNamespace("sommer", quietly=TRUE) &&
   "blasThreads" %in% names(formals(sommer::mmes)) &&
   requireNamespace("parallelly", quietly=TRUE) &&
   (hasFlexiBLAS || hasDirectControl)){
  local({
    suppressPackageStartupMessages(library(sommer))
    getThreads <- if(hasFlexiBLAS) flexiblas::flexiblas_get_num_threads else
      RhpcBLASctl::blas_get_num_procs
    before <- getThreads()
    exampleData <- expand.grid(Name=factor(seq_len(12L)),
      Env=factor(letters[1:3]), replicate=seq_len(2L))
    exampleData$Yield <- as.integer(exampleData$Env) +
      sin(as.integer(exampleData$Name)) + cos(seq_len(nrow(exampleData)))
    for(request in list(1L, "auto")){
      fit <- sommer::mmes(Yield~Env, random=~Name, rcov=~units,
        data=exampleData, solver="cholmod", nIters=2,
        dateWarning=FALSE, verbose=FALSE, blasThreads=request)
      print(fit$threadingDiagnostics[c("blasBackend", "controller",
        "blasThreadsRequested", "blasThreadsConfigured", "availableCPUs",
        "openmpMaxThreads", "numericalPath", "averageCPUs")])
      stopifnot(getThreads() == before)
    }
  })
}else{
  message("Verification skipped: install the matching controller and parallelly, and load a sommer build with blasThreads.")
}
```

`blasThreadsConfigured` should show the setting used during the fit. The
controller restores the previous BLAS setting on return, including on errors.
Therefore, querying BLAS threads *after* `mmes()` returns can show the old
count; that does not mean the request was ignored. Use the returned diagnostics
to inspect the fit's configuration.

# Retry your model

After completing the route matching your R installation, the original model
can be fitted without changing its statistical specification:

```{r user-model, eval=FALSE}
library(sommer)
mixm2 <- mmes(value~envf, nIters=5,
  random=~vsm(fam(envf, 1), ism(id), Gu=MMTi) +
           vsm(dsm(envf), ism(repf)),
  rcov=~vsm(dsm(envf), ism(units)),
  data=DT, blasThreads="auto")

mixm2$threadingDiagnostics
```

Here `DT` and `MMTi` are your existing data and inverse relationship matrix.
Alternatively, request an explicit count such as `blasThreads=4L`. Installing
`parallelly` alone only resolves automatic CPU detection; it does not supply
BLAS control.

# Cluster allocations and startup configuration

`"auto"` uses `parallelly::availableCores()` rather than blindly using every
CPU visible on the host. Scheduler limits, affinity and container quotas can
make this count smaller than `parallel::detectCores()`. It does not request
additional CPUs from a scheduler or distribute a fit across nodes.

For example, a Slurm job requesting eight CPUs for one R process can start with:

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8

# Load your site's compatible R/numerical-library modules here.
export OMP_NUM_THREADS=1
Rscript fit.R
```

With a working runtime controller, `fit.R` can use `blasThreads="auto"`.
Setting OpenMP to one isolates BLAS threading and avoids many OpenMP workers
each launching a BLAS team; it is a starting point to benchmark, not a
universally optimal setting. BLAS and Sommer's OpenMP loops are separate
threading layers. Some BLAS builds themselves use OpenMP, so startup variables
and runtime setters can interact.

If runtime control is unavailable but your BLAS supports startup thread
configuration, leave `blasThreads=NULL` and configure it **before starting R**.
For a pthread-based OpenBLAS backend, for example:

```bash
OPENBLAS_NUM_THREADS=8 OMP_NUM_THREADS=1 Rscript fit.R
```

For MKL, use its documented `MKL_NUM_THREADS` setting; OpenMP-based OpenBLAS
uses `OMP_NUM_THREADS`. Follow the selected backend's documentation. These
variables cannot make a serial reference BLAS threaded, and changing them
with `Sys.setenv()` after BLAS initializes may have no effect. Without a
supported query interface, the BLAS thread count can remain unknown to Sommer
even when startup configuration works.

# Troubleshooting and interpreting usage

| Symptom | Check or action |
|---|---|
| `Runtime BLAS control is unavailable` | Check the BLAS identity in this R session; install its matching adapter in an active `.libPaths()` library. |
| `blasThreads='auto' requires ... parallelly` | Install `parallelly`, or supply a positive integer instead. |
| `flexiblas_avail()` is `FALSE` | Use R linked to FlexiBLAS; installing its R interface does not relink R. |
| Adapter installed but BLAS has a generic name | Current detection may not recognize that installation; use startup settings with `blasThreads=NULL` or an explicitly identified supported R build. |
| Backend did not accept the requested setting | Check whether it is serial or has a thread limit; select a supported threaded backend or use an accepted count. |
| `unused argument (blasThreads=...)` | The loaded Sommer is older than the source tree; install the current tree and restart R. |
| Packages available interactively but missing in a job | Check the job's R module, `.libPaths()` and `R_LIBS_USER`; install or expose packages for that R installation. |

Configured counts are not actual utilization. `openmpMaxThreads` is an OpenMP
limit, not a CHOLMOD or BLAS usage count. `averageCPUs` measures aggregate
process CPU seconds divided by elapsed seconds, not an instantaneous number
of active threads. Factorization measurements are in
`threadingDiagnostics$factorizations`; their `measurement` fields distinguish
numeric CHOLMOD/LDLT work from the broader block-Schur workload.

Even `solver="cholmod"` can select a structure-aware Schur engine for an
eligible model. Small blocks, serial work, memory bandwidth and thread startup
overhead can prevent speedups. Confirm the numerical path and benchmark several
counts on representative data rather than assuming that `"auto"` is fastest.
The repository's `benchmarks/thread-scaling.R` and accompanying README provide
a repeated scaling benchmark with numerical-agreement checks.

# References

- [flexiblas package documentation](https://cran.r-project.org/package=flexiblas)
- [RhpcBLASctl package documentation](https://cran.r-project.org/package=RhpcBLASctl)
- [parallelly package documentation](https://cran.r-project.org/package=parallelly)
- [R Installation and Administration](https://cran.r-project.org/doc/manuals/r-release/R-admin.html)